---
title: "EEMB R Seminar: ggplot"
author: "Indra Boving"
date: "1/12/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE) #setup for markdown document
```

```{r}
library(tidyverse) 
library(here) #Easy to find data!
library(lubridate) #Dealing with dates. 
library(janitor) #Cleaning up headings, etc. 
library(ggbeeswarm) #specific point aesthetics for ggplot
library(dlookr) #for using diagnose()
```

#Trait data: 

Data source: 

Law, B. E., & Berner, L. T. (2015, July 20). NACP TERRA-PNW: Forest Plant Traits, NPP, Biomass, Soil Properties.
accessible here: https://doi.org/10.3334/ORNLDAAC/1292

full data description is available here:

Berner, L. T., & Law, B. E. (2016). Plant traits, productivity, biomass and soil properties from forest sites in the Pacific Northwest, 1999–2014. Scientific Data, 3, 160002. https://doi.org/10.1038/sdata.2016.2

Read in data: 
```{r}
trait_data <- read.csv(here("data", "PNW_data_for_Anderegg_2018_clean.csv")) %>% #use here() to tell read.csv where to look for the file (can also use here() with write.csv to organize where to put files)
  clean_names() %>% #make all column titles lower_snake_case
  mutate(name = binomial) #rename columns 

dlookr::diagnose(trait_data) #take a look at the data
```


### Scatterplot: 

Make a basic scatterplot: 
```{r}
scatter_plain <- trait_data %>% #pipe data into the plot
  ggplot(aes(y = leaf_nitrogen, x = leaf_life)) + #show aesthetics
  geom_point(aes(color = name)) #color by a group, in this case species: 
scatter_plain
```
...overwhelming! Maybe we aren't interested in the specific species, so let's color by something more broad:

```{r}
trait_data %>% #pipe data into the plot
  ggplot(aes(y = leaf_nitrogen, x = leaf_life)) + #show aesthetics
  geom_point(aes(color = pft)) #color by a group
```
Let's say we want to look at just one ecoregion: 

```{r}
e_cascades <- trait_data %>% #just look at a subset of the data
  filter(ecoregion == "Eastern Cascades") #from a specific ecoregion...
```

```{r}
scatter_plain <- e_cascades %>% #pipe data into the plot
  ggplot(aes(y = leaf_nitrogen, x = leaf_life)) + #show aesthetics
  geom_point(aes(color = name)) #color by a group
scatter_plain
```
Update the plot in a few ways: 
```{r}
scatter_labeled <- scatter_plain + #take older plot...
  theme_minimal() + #...and change the theme
  scale_x_continuous(breaks = c(4, 8, 12, 16)) + #break up the x-axis to smaller increments
  scale_color_brewer(palette = "Dark2") + #change colors 
  labs(x = "Leaf Lifespan (Years)", #add labels
       y = "Leaf Nitrogen (%)", 
       color = "Species") 
scatter_labeled
```
...could also change shape to highlight the deciduous group: 
```{r}
scatter_pft <- e_cascades %>% 
  ggplot(aes(y = leaf_nitrogen, x = leaf_life)) + 
  geom_point(aes(color = name, 
                 shape = pft)) + #add shape to pft
  theme_minimal() + 
  scale_x_continuous(breaks = c(4, 8, 12, 16)) + 
  scale_color_brewer(palette = "Dark2") +  
  labs(x = "Leaf Lifespan (Years)", 
       y = "Leaf Nitrogen (%)", 
       color = "Species", 
       shape = "Functional Type") #update labels 
scatter_pft
```

Add annotation: 
```{r}
scatter_annotated <- scatter_pft +
  annotate("segment", x = 6, xend = 1.5, y = 2.5, yend = 2, #tell annotate() where the segment should go
           colour = "black", size = .5, 
           arrow = arrow(ends = "last", angle = 25, type = "closed", length = unit(.1,"cm")))  + #tell it that you want it to be an arrow
  annotate("text", x= 6, y= 2.59, label= "Deciduous", size = 4) #add a text box
scatter_annotated
```

### Boxplots:

Basic boxplot:
```{r, echo = FALSE, warning=FALSE}
e_cascades %>% 
  ggplot(aes(y = specific_leaf_area, x = family)) + #specify ggplot aesthetics
  geom_boxplot()
```
A nicer way to visualize the same data: 

```{r}
e_cascades %>% 
  ggplot(aes(y = specific_leaf_area, x = family)) + #specify ggplot aesthetics
  geom_beeswarm(aes(color = family), 
                cex = 1, #point spacing
                size = 1,  #point size
                alpha = .7) + #color the points, use beeswarm +
  geom_boxplot(fill = NA, #layer on a boxplot...
               width = 0.5, 
               outlier.color = NA) + #...that has no fill
  stat_summary(fun=mean, #specify what statistics we want to be showing with our boxplots
               geom="point", #...in this case, we want to show the mean represented by a black point
               shape=20, 
               size=4, 
               color="black", 
               fill="black") + 
  labs(y = "SLA", #replace labels
      x = "Family", 
      title = "East Cascades") +
  scale_color_manual(values = c("olivedrab", "lightblue", "darkblue")) + #specify colors manually 
  theme_minimal() + #give it a theme
  theme(legend.position = "none") #remove the legend
```

Step by step:

```{r, echo = FALSE, warning=FALSE}
#make first plot 
box_beeswarm <- e_cascades %>% 
  ggplot(aes(y = specific_leaf_area, x = family)) + #specify ggplot aesthetics
  #geom_point(aes(color = family)) +
  geom_beeswarm(aes(color = family), 
                cex = 1, #point spacing
                size = 1,  #point size
                alpha = .7)  #color the points, use beeswarm
box_beeswarm
```


```{r, echo = FALSE, warning=FALSE}
#layer on second plot
beeswarm_boxplot <- box_beeswarm +
  geom_boxplot(fill = NA, #layer on a boxplot...
               width = 0.5, 
               outlier.color = NA) + #...that has no fill
  stat_summary(fun=mean, #specify what statistics we want to be showing with our boxplots
               geom="point", #...in this case, we want to show the mean represented by a black point
               shape=20, 
               size=4, 
               color="black", 
               fill="black")
beeswarm_boxplot
```


```{r, echo = FALSE, warning=FALSE}
#add labels and change colors and theme: 
boxplot_labeled<- beeswarm_boxplot + #take older figure
  labs(y = "SLA", #replace labels
      x = "Family", 
      title = "East Cascades") +
  scale_color_manual(values = c("olivedrab", "lightblue", "darkblue")) + #specify colors manually 
  theme_minimal() + #give it a theme
  theme(legend.position = "none") #remove the legend

boxplot_labeled
```


# Chaparral data: 

Data source: 

Bartosh, H., B. Peterson, M.M. Halbur, T.L. Comendant, A.L. Whipple, and E. Micheli. 2020. Post-Tubbs Fire Chaparral Floristic Survey at Pepperwood Preserve in the California Coast Ranges 2018-2019 ver 1. Environmental Data Initiative. https://doi.org/10.6073/pasta/958a452d488a54100965e99ecda28fd8 (Accessed 2022-01-19).

Great source for other environmental data: https://portal.edirepository.org


```{r}
chaparral_data_raw <- read.csv(here("data", "chaparral_pepperwood.csv")) #read in data

dlookr::diagnose(chaparral_data_raw) #check out data
```

Use lubridate to make dates easier to work with: 

```{r}
chaparral_data_clean <- chaparral_data_raw %>% 
  clean_names() %>% #make all columns lower_snake_case
  mutate(date = ymd(date_collected), #use lubridate package to convert date column into date format
         year = year(date), #use lubridate to extract year
         month = month(date)) %>% #use lubridate to extract month
  filter(plant_type %in% c("sprout", "seed")) #filter out only sprouters and seeders

#unique(chaparral_data_raw$timing) #view how many unique values in a column
class(chaparral_data_clean$date)
```

### Bar plots:

```{r, echo = FALSE, warning=FALSE}
chaparral_data_clean %>% 
  ggplot(aes(y = percent_cover, x = fire_severity, fill = plant_type)) + #specify ggplot aesthetics
  geom_col(position = "dodge") +
  facet_wrap(~plant_type) + #we want to compare plant types and their percent cover post-fire
  theme_light() + 
  theme(legend.position = "none")
```

Rename some columns and reorder some factors so they appear in the correct order:

```{r, echo = FALSE, warning=FALSE}
chaparral_data_clean <- chaparral_data_clean %>% 
  mutate(fire_severity = recode(fire_severity, "med" = "Medium", "high" = "High", "low" = "Low")) %>%  #change labels to be upper case
  mutate(fire_severity = fct_relevel(fire_severity, "Low", "Medium", "High")) %>% #order those levels so they are in a logical order
  mutate(plant_type = recode(plant_type, "sprout" = "Sprout", "seed" = "Seed"))
```

One way to visualize side-by-side columns: 

```{r, echo = FALSE, warning=FALSE}
chaparral_data_clean %>% 
  ggplot(aes(y = percent_cover, x = fire_severity, fill = plant_type)) + #specify ggplot aesthetics
  geom_col(position = "dodge") +
  facet_wrap(~plant_type) + #we want to compare plant types and their percent cover post-fire
  labs(y = "Percent Cover (%)", 
      x = "Fire Severity", 
      title = "Effect of fire severity on percent cover") +
  scale_fill_manual(values = c("olivedrab", "lightblue", "darkblue")) +
  theme_light() + 
  theme(legend.position = "none")
```

Another way to visualize side-by-side columns:  

```{r, echo = FALSE, warning=FALSE}
chaparral_data_clean %>% 
  ggplot(aes(y = percent_cover, x = fire_severity, fill = plant_type)) + #specify ggplot aesthetics
  geom_col(position = "dodge") +
 # facet_wrap(~plant_type) + #we want to compare plant types and their percent cover post-fire
  labs(y = "Percent Cover (%)", 
      x = "Fire Severity", 
      title = "Effect of fire severity on percent cover") +
  scale_fill_manual(values = c("olivedrab", "lightblue", "darkblue")) +
  theme_light() + 
  theme(legend.position = "none")
```


